\Title

### 2023-02-10 Grimaldi-etal-BiolCybernetics.tex
methods : 2.2

%In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study. Please refer to suggested Data Availability Statements in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}. If the study did not report any data, you might add ``Not applicable'' here. Chen
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% NOTATIONS
% Pixel world
\newcommand{\presynaddr}{a} % pre address
\newcommand{\postsynaddr}{b} % post address
\newcommand{\numevent}{N_{ev}} % total number of events
\newcommand{\presynaddrspace}{\mathcal{A}} %presynaptic address space
\newcommand{\postsynaddrspace}{\mathcal{B}} %postsynaptic address space
\newcommand{\Npol}{N_\text{p}} % number of polarity
\newcommand{\Nneuron}{N_\text{n}} % number of output neurons in the layer
\newcommand{\arank}{r} % address index
\newcommand{\bias}{\beta} % bias for the MLR model
\newcommand{\synapse}{\mathcal{S}} % synapse
\newcommand{\synapticweight}{w} % synaptic weight
\newcommand{\synapticdelay}{\delta} % synaptic delay
\newcommand{\ranksyn}{s} % synapse index
\newcommand{\Nsyn}{N_{s}} % total number of synapses
\newcommand{\activeweights}{\mathcal{W}} 
\newcommand{\timev}{t} % time
\newcommand{\polev}{p} % polarity
\newcommand{\event}{\epsilon} % event
\newcommand{\eventstream}{\xi} % stream of events
\newcommand{\TS}{S} % time surface
\newcommand{\neuron}{\mathbf{n}} % neuron in the SNN (defined by the spatial position and the channel)
\newcommand{\postneuron}{\mathbf{m}} % post synaptic neuron in the SNN (defined by the spatial position and the kernel)
\newcommand{\channel}{\mathbf{p}} % channel
\newcommand{\layer}{\mathbf{L}} % layer
\newcommand{\ms}{\si{\milli\second}}%
\newcommand{\us}{\si{\micro\second}}%
\newcommand{\timecontext}{T} % time context (cf HOTS) matrice gathering last event times
\newcommand{\current}{I} % post synaptic current
\newcommand{\volt}{u} % membrane potential
\newcommand{\volts}{V} % matrix of membrane potentials
\newcommand{\gain}{\gamma} % homeostatic gain
\newcommand{\simil}{\beta} % similarity value
\newcommand{\Nclass}{N_\text{class}} % number of classes for MLR:
\newcommand{\Nx}{N_\text{X}}
\newcommand{\Ny}{N_\text{Y}}
\newcommand{\Ntime}{N_\text{T}}
\newcommand{\kernel}{\mathcal{K}} % convolution kernel
%\newcommand{\kernelind}{\mathbf{k}} % indice of the kernel
\newcommand{\kernelind}{k} % indice of the kernel
\newcommand{\Kx}{K_\text{X}}
\newcommand{\Ky}{K_\text{Y}}
\newcommand{\Ktime}{K_\text{T}}
\newcommand{\classiflayer}{\mathbf{C}}
\newcommand{\class}{c} % class k of the MLR
\newcommand{\lrweights}{\theta} % matrix of MLR weights
\newcommand{\lrtrue}{y} % true value of the prediction for MLR
\newcommand{\loss}{J} % cost function for MLR
\newcommand{\softmax}{\sigma}
\newcommand{\actfreq}{f}
\newcommand{\decision}{\hat{y}}
\newcommand{\colorsec}{black}
\newcommand{\colorsubsec}{black}
\newcommand{\speed}{v}
\newcommand{\Nspeed}{N_v}
% Example definitions.
% --------------------
\def\x{{\mathbf x}}
\def\L{{\cal L}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Of interest is to define the receptive field of a postsynaptic neuron $\synapse^\postsynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \postsynaddr_\ranksyn=\postsynaddr\}_{\ranksyn \in [1,\Nsyn]}  \subset \synapse$, or the emitting field of a presynaptic neuron $\synapse_\presynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \presynaddr_\ranksyn=\presynaddr\}_{\ranksyn \in [1,\Nsyn]} \subset \synapse$. As a consequence, a postsynaptic neuron $\postsynaddr$ receives an event stream which is multiplexed by the synapses of its receptive field. It results in a weighted event stream for each  postsynaptic neuron $\postsynaddr$: 
\begin{equation}\label{eq:stream_b}
\event_\postsynaddr = \{(\presynaddr_\arank, \synapticweight_\arank, \timev_\arank+\synapticdelay_\ranksyn) \| \presynaddr_\arank = \presynaddr_\ranksyn \}_{\arank \in [1,\numevent], \ranksyn \in \synapse^\postsynaddr}
\end{equation}
% Unlike the binary stream of events from the signal as input, the postsynaptic neurons receive a weighted stream of event such that for each event $\arank$ its value is $\synapticweight_\arank = \synapticweight_\ranksyn$ when $\presynaddr_\arank = \presynaddr_\ranksyn$.
This new stream of events is ordered in time in biology as events reach the soma of post-synaptic neurons but should be reordered in simulations. Crucially, when post-synaptic neurons are activated on their soma by a specific spatio-temporal motif which is imprinted in the set of synapses, the discharge probability will increase, notably when these spikes converge on the soma in a synchronous manner while the activation function of the neurons of the HD-SNN can be selected among the whole range of spiking neuron response functions (see Figure~\ref{fig:izhikevich}). We defined the HD-SNN model in this subsection in all generality and in the next subsection we describe an implementation of our model adapted for the motion detection task. 
%
% \note{I can make a small figure with event streams of different colors for the input one, the multiplexed one etc and the response in the bottom}
%
\subsection{Application of HD-SNN to the motion detection task}
%
From the perspective of simulating such event-based computations on standard CPU- or GPU-based computers and then using parallel computing, it is useful to transform this event-based representation into a dense representation. Indeed, by discretization of time, we may transform any event-based input as a boolean matrix $A \in \{0, 1 \}^{\Npol \times \Ntime \times \Nx \times \Ny}$, defined for all polarities $p$, times $t$ and space coordinates $x$ and $y$. The values are by definition equal to zero except at the occurences of events: $\forall \numevent \in \mathbb{N}$, $A()=1$

$\event = \{(\presynaddr_\arank, \timev_\arank)\}_{\arank \in [1,\numevent]}$ where $\numevent \in \mathbb{N} $ is the total number of events in the data stream and the rank $\arank$ is the index of each event in the list of events. Events are typically ordered by their time of occurrence. Each event has a time of occurrence $\timev_\arank$ and an associated address $\presynaddr_\arank$ (which is typically in the form $(x_\arank, y_\arank, \polev_\arank)$ for event-based cameras).

To benefit from invariance to location observed in CNN and to simplify the implementation of the model, one assumes that kernels should be similar across different positions and define a convolution operator. % The characteristics of the kernels are shared among one class of neurons such that $\forall i,j \in [1, N_\class]^2, \; \kernel^i_\class = \kernel^j_\class$. 
% Here, we will consider that heterogeneous delays are limited in range
In particular, the whole synaptic set can be represented as a one kernel for each class $\class$ of the supervision task as the dense matrix $\kernel_\class$ of size $(\Npol, \Ktime, \Kx, \Ky)$, where  $\Ktime$ is the number of delays and $\Kx$ and $\Ky$ are the number of pixels in both spatial dimensions. To keep an analogy with the HD-SNN model, for each neuron $\postsynaddr$ of position $(x, y)$ and class $\class$, $\kernel_\class$ gives the weights as a function of presynaptic address and delay: $\forall {\ranksyn \in \synapse^\postsynaddr}, \kernel_\class(\presynaddr_\ranksyn, \synapticdelay_\ranksyn) = \synapticweight_\ranksyn$. Therefore, if we make the approximation that one neuron integrates only on the temporal window given by its variety of synaptic delays, the integration of the spike train as input can be formalized by a 3D convolution operation. The longest synaptic delay defines the depth of the kernel $\kernel^\postsynaddr$ and all possible delays associated to the different presynaptic addresses are represented. Using this dense representation, the processing of $A$ by the model can be written as:
\begin{equation}\label{eq:kernel_b}
\forall x, y, \timev, \; \;
\mathcal{C}_\class(x, y, \timev)
= \sum_{p, \delta_x, \delta_y, \synapticdelay} \kernel_\class(p, \delta_x, \delta_y, \synapticdelay) \cdot A(p, x - \delta_x, y - \delta_y, \timev-\synapticdelay)
\end{equation}
% The computational brick used in equation~\eqref{eq:kernel_b} can be extended to every position, in space and time, of the boolean matrix $A$:
% \begin{equation}\label{eq:conv}
% \forall \presynaddr \in \presynaddrspace, \forall \timev \in T, \; \;
% \mathcal{C}_\class(\presynaddr,\timev)
% = \sum_{\ranksyn  \in [1, N_{\synapse^\postsynaddr}]} \kernel_\class(\presynaddr_\ranksyn, \synapticdelay_\ranksyn) \cdot A(\presynaddr, \timev-\synapticdelay_\ranksyn)
% \end{equation}
This shows that $\mathcal{C}_\class$ is a spatiotemporal convolution of the dense representation of the event stream with the dense kernels formed by the set of synapses:  $\mathcal{C}_\class = \kernel_\class \ast A$ (see Figure~\ref{fig:model} for an illustration). Note that to remain within the framework of a causal calculation, the kernels are shifted in time such that only past information gives an answer at the present time.
This well-known computation defines a differentiable measure which is very efficiently implemented for GPUs and which we will use for learning the classification of different motifs in the event stream.

The same type of spatiotemporal filtering is used as a preprocessing stage of a pattern recognition algorithm~\cite{ghosh_spatiotemporal_2019}. And in \cite{sekikawa_constant_2018}, an efficient 3D convolutional algorithm succeeds a motion estimation task. By assuming a locally constant velocity, the 3D kernel can be decomposed into a 2D kernel representing the shapes and a 3D kernel representing the velocity. Here, we keep the analogy with spiking neurons, and we try to use as few heuristics as possible to observe the emergence of spiking motifs in the spatiotemporal kernels. If so, this precise spatiotemporal patterns prove to be of interest for neural computations and one can assume that biological neurons make use of this information as well. 
% \note{plus de biblio sur la motion detection/segmentation ou optique flow avec les events? (reviewer 2)}
The 3D convolution represents the linear integration of the spike train as input. The activation function of our model is a softmax function implementing a form of Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. In our MLR model, a probability value, for each class (i.e. each motion direction), is predicted for each event at position $x, y$ and at time $\timev$ as a softmax function of the linear combination of the list of events. % on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the HD-SNN model. 
Using the kernels, it transforms the input raster plot into a probability with the following formula:
\begin{equation}\label{eq:mlr}
    \forall x, y, \timev, \forall \class \in [1, N_\class], \; \;
Pr(k=\class \; \vert \;  x, y, \timev) =
\frac {\exp  (\mathcal{C}_\class(x, y, \timev) +\bias_\class) }{Z(x, y, \timev)}
\end{equation} 
where $Z(x, y, \timev)=\sum_{\class \in [1, N_\class]} \exp  (\mathcal{C}_\class(x, y, \timev) +\bias_\class)$ is the partition function and  $\bias_\class$ is the bias linked to class $\class$. 
In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike. The spiking output of the model corresponds to an event with the highest probability class.

@@

We would like to highlight a few innovations in the contributions that are presented in this paper. First, whereas~\citep{ghosh_spatiotemporal_2019,yu_stsc-snn_2022} use a correlation-based heuristic, which we observed to be less efficient, the generic heterogeneous model is formalized from first principles for optimal detection of the event-based spatio-temporal motifs. Moreover, in comparison to HOTS~\citep{lagorce_hots_2017}, the weights are explicable as they directly inform on the logit (inverse sigmoid of the probability) of detection of each spatio-temporal spike motif. Another novelty is that the model learns the weights and the delays simultaneously. For example, the polychronization model~\citep{izhikevich_polychronization_2006} learns only the weights using STDP, while the delays are randomly drawn and their values are frozen during learning. In addition, the model is evaluated on a realistic task, while models such as the tempotron are tested on simplified toy problems~\citep{gutig_tempotron_2006}. Another major contribution is to provide a model that is suitable for learning any kind of spatio-temporal spiking motifs and that can be trained in a supervised manner by providing a dataset of supervision pairs. Instead of relying on a careful description of the physical rules governing a task, e.g. the luminance conservation principle for motion detection~\citep{benosman_asynchronous_2012, dardelet_event-by-event_2021}, this allows a more flexible definition of the model using this properly labeled dataset.
%
\subsection{Main limits}
%%%-----------------------------------------------------------------
We have identified a number of limitations of our model, which we will now discuss in detail. First, the entire framework is based on a discrete binning of time, which is not compatible with the continuous nature of biological time. We used this binning to efficiently implement the framework on conventional hardware, especially GPUs, to be able to use fast three-dimensional convolutions. We have tested the effect of the width of the time bin and shown that it has essentially no effect on the results presented in this paper. This is consistent with the relative robustness of other event-based frameworks such as HOTS~\citep{lagorce_hots_2017}, where accuracy was unaffected when the input spikes were subjected to noisy perturbations up to $1~\ms$~\citep{grimaldi_robust_2022}. This suggests the possibility of analytically including a precision term in the temporal value of the input spikes. This mechanism may be implemented by the filtering implemented by the synaptic time constant of about $5~\ms$. Furthermore, it is possible to circumvent the need for time discretization by the use of a purely event-based scheme. In fact, it is not necessary to compute the voltage traces between two spikes~\citep{hanuschkin_general_2010}. Thus, it is possible to define a purely event-based framework. Such an architecture could provide promising computational speedups. % \note{pourrait parler du poster bernstein? de nadafian?}

>>>>

%In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study. Please refer to suggested Data Availability Statements in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}. If the study did not report any data, you might add ``Not applicable'' here. Chen
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% NOTATIONS
\newcommand{\ms}{\si{\milli\second}}%
\newcommand{\us}{\si{\micro\second}}%
\newcommand{\presynaddr}{a} % pre address
\newcommand{\postsynaddr}{b} % post address
\newcommand{\numevent}{N_{ev}} % total number of events
\newcommand{\presynaddrspace}{\mathcal{A}} %presynaptic address space
\newcommand{\postsynaddrspace}{\mathcal{B}} %postsynaptic address space
\newcommand{\Npol}{N_\text{p}} % number of polarity
\newcommand{\Nneuron}{N_\text{n}} % number of output neurons in the layer
\newcommand{\arank}{r} % address index
\newcommand{\bias}{\beta} % bias for the MLR model
\newcommand{\synapse}{\mathcal{S}} % synapse
\newcommand{\synapticweight}{w} % synaptic weight
\newcommand{\synapticdelay}{\delta} % synaptic delay
\newcommand{\ranksyn}{s} % synapse index
\newcommand{\Nsyn}{N_{s}} % total number of synapses
\newcommand{\activeweights}{\mathcal{W}} 
\newcommand{\timev}{t} % time
\newcommand{\polev}{p} % polarity
\newcommand{\event}{\epsilon} % event
\newcommand{\eventstream}{\xi} % stream of events
\newcommand{\TS}{S} % time surface
\newcommand{\neuron}{\mathbf{n}} % neuron in the SNN (defined by the spatial position and the channel)
\newcommand{\postneuron}{\mathbf{m}} % post synaptic neuron in the SNN (defined by the spatial position and the kernel)
\newcommand{\channel}{\mathbf{p}} % channel
\newcommand{\layer}{\mathbf{L}} % layer
% \newcommand{\timecontext}{T} % time context (cf HOTS) matrice gathering last event times
\newcommand{\current}{I} % post synaptic current
\newcommand{\volt}{u} % membrane potential
\newcommand{\volts}{V} % matrix of membrane potentials
\newcommand{\gain}{\gamma} % homeostatic gain
% \newcommand{\simil}{\beta} % similarity value
\newcommand{\Nclass}{N_\text{class}} % number of classes for MLR:
\newcommand{\Nx}{N_\text{X}}
\newcommand{\Ny}{N_\text{Y}}
\newcommand{\Ntime}{N_\text{T}}
\newcommand{\kernel}{\mathcal{K}} % convolution kernel
%\newcommand{\kernelind}{\mathbf{k}} % indice of the kernel
\newcommand{\kernelind}{k} % indice of the kernel
\newcommand{\Kx}{K_\text{X}}
\newcommand{\Ky}{K_\text{Y}}
\newcommand{\Ktime}{K_\text{T}}
\newcommand{\classiflayer}{\mathbf{C}}
\newcommand{\class}{c} % class k of the MLR
\newcommand{\lrweights}{\theta} % matrix of MLR weights
\newcommand{\lrtrue}{y} % true value of the prediction for MLR
\newcommand{\loss}{J} % cost function for MLR
\newcommand{\softmax}{\sigma}
\newcommand{\actfreq}{f}
\newcommand{\decision}{\hat{y}}
\newcommand{\colorsec}{black}
\newcommand{\colorsubsec}{black}
\newcommand{\speed}{v}
\newcommand{\Nspeed}{N_v}
% Example definitions.
% --------------------
\def\x{{\mathbf x}}
\def\L{{\cal L}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Of interest is to define the receptive field of a postsynaptic neuron $\synapse^\postsynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \postsynaddr_\ranksyn=\postsynaddr\}_{\ranksyn \in [1,\Nsyn]}  \subset \synapse$, or the emitting field of a presynaptic neuron $\synapse_\presynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \presynaddr_\ranksyn=\presynaddr\}_{\ranksyn \in [1,\Nsyn]} \subset \synapse$. As a consequence, a postsynaptic neuron $\postsynaddr$ receives an event stream which is multiplexed by the synapses of its receptive field. It results in a weighted event stream for each  postsynaptic neuron $\postsynaddr$: 
\begin{equation}\label{eq:stream_b}
\event_\postsynaddr = \{(\presynaddr_\arank, \synapticweight_\arank, \timev_\arank+\synapticdelay_\ranksyn) \| \presynaddr_\arank = \presynaddr_\ranksyn \}_{\arank \in [1,\numevent], \ranksyn \in \synapse^\postsynaddr}
\end{equation}
% Unlike the binary stream of events from the signal as input, the postsynaptic neurons receive a weighted stream of event such that for each event $\arank$ its value is $\synapticweight_\arank = \synapticweight_\ranksyn$ when $\presynaddr_\arank = \presynaddr_\ranksyn$.
This new stream of events is ordered in time in biology as events reach the soma of post-synaptic neurons but should be reordered in simulations. Crucially, when post-synaptic neurons are activated on their soma by a specific spatio-temporal motif which is imprinted in the set of synapses, the discharge probability will increase, notably when these spikes converge on the soma in a synchronous manner while the activation function of the neurons of the HD-SNN can be selected among the whole range of spiking neuron response functions (see Figure~\ref{fig:izhikevich}). We defined the HD-SNN model in this subsection in all generality and in the next subsection we describe an implementation of our model adapted for the motion detection task. 
% \note{I can make a small figure with event streams of different colors for the input one, the multiplexed one etc and the response in the bottom}
\subsection{Application of HD-SNN to the motion detection task}
From the perspective of simulating such event-based computations on standard CPU- or GPU-based computers and then using parallel computing, it is useful to transform this event-based representation into a dense representation. Indeed, by discretization of time, we may transform any event-based input as a boolean matrix $A \in \{0, 1 \}^{\Npol \times \Ntime \times \Nx \times \Ny}$, defined for all polarities $p$, times $t$ and space coordinates $x$ and $y$. The values are by definition equal to zero except at the occurrences of events: $\forall \numevent \in \mathbb{N}$, $A(\polev_\arank, t_\arank, x_\arank, y_\arank)=1$. First, one can note that by using a discretization, the computational brick used in equation~\eqref{eq:kernel_b} corresponds to a temporal convolution which transforms the input $A$ using one kernel per postsynaptic neuron~\citep{grimaldi_learning_2022}. To benefit from invariance to location observed in images and exploited in Convolutional Neural Networks, one may further assume that synaptic motifs should be similar across different positions such that we may define a spatio-temporal convolution operator. % The characteristics of the kernels are shared among one class of neurons such that $\forall i,j \in [1, N_\class]^2, \; \kernel^i_\class = \kernel^j_\class$. 
% Here, we will consider that heterogeneous delays are limited in range
In particular, the whole synaptic set can be represented as a one kernel for each class $\class$ of the supervision task as the dense matrix $\kernel_\class$ of size $(\Npol, \Ktime, \Kx, \Ky)$, where  $\Ktime$ is the number of delays and $\Kx$ and $\Ky$ are the number of pixels in both spatial dimensions. To keep an analogy with the HD-SNN model, for each neuron $\postsynaddr$ of position $(x, y)$ and class $\class$, $\kernel_\class$ gives the weights as a function of presynaptic address and delay: $\forall {\ranksyn \in \synapse^\postsynaddr}, \kernel_\class(\presynaddr_\ranksyn, \synapticdelay_\ranksyn) = \synapticweight_\ranksyn$. %Therefore, if we make the approximation that one neuron integrates only on the temporal window given by its variety of synaptic delays, the integration of the spike train as input can be formalized by a 3D convolution operation. The longest synaptic delay defines the depth of the kernel $\kernel^\postsynaddr$ and all possible delays associated to the different presynaptic addresses are represented. 
Using this dense representation, the processing of $A$ by the model can be written as:
\begin{equation}\label{eq:kernel_b}
\forall x, y, \timev, \; \;
\mathcal{C}_\class(x, y, \timev)
= \sum_{p, \delta_x, \delta_y, \synapticdelay} \kernel_\class(p, \delta_x, \delta_y, \synapticdelay) \cdot A(p, x - \delta_x, y - \delta_y, \timev-\synapticdelay)
\end{equation}
% The computational brick used in equation~\eqref{eq:kernel_b} can be extended to every position, in space and time, of the boolean matrix $A$: equation~\eqref{eq:stream_b}
% \begin{equation}\label{eq:conv}
% \forall \presynaddr \in \presynaddrspace, \forall \timev \in T, \; \;
% \mathcal{C}_\class(\presynaddr,\timev)
% = \sum_{\ranksyn  \in [1, N_{\synapse^\postsynaddr}]} \kernel_\class(\presynaddr_\ranksyn, \synapticdelay_\ranksyn) \cdot A(\presynaddr, \timev-\synapticdelay_\ranksyn)
% \end{equation}
This shows that $\mathcal{C}_\class$ is a spatio-temporal convolution of the dense representation of the event stream with the dense kernels formed by the set of synapses:  $\mathcal{C}_\class = \kernel_\class \ast A$ (see Figure~\ref{fig:model} for an illustration). Note that to remain within the framework of a causal calculation, the kernels are shifted in time such that only past information gives an answer at the present time.
This well-known computation defines a differentiable measure which is very efficiently implemented for GPUs and which we will use for learning the classification of different motifs in the event stream.

The same type of spatio-temporal filtering is used as a preprocessing stage of a pattern recognition algorithm~\cite{ghosh_spatio-temporal_2019}. And in \cite{sekikawa_constant_2018}, an efficient 3D convolutional algorithm succeeds a motion estimation task. By assuming a locally constant velocity, the 3D kernel can be decomposed into a 2D kernel representing the shapes and a 3D kernel representing the velocity. Here, we keep the analogy with spiking neurons, and we try to use as few heuristics as possible to observe the emergence of spiking motifs in the spatio-temporal kernels. If so, this precise spatio-temporal patterns prove to be of interest for neural computations and one can assume that biological neurons make use of this information as well. 
% \note{plus de biblio sur la motion detection/segmentation ou optique flow avec les events? (reviewer 2)}
The 3D convolution represents the linear integration of the spike train as input. The activation function of our model is a softmax function implementing a form of Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. In our MLR model, a probability value, for each class (i.e. each motion direction), is predicted for each event at position $x, y$ and at time $\timev$ as a softmax function of the linear combination of the list of events. % on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the HD-SNN model. 
Using the kernels, it transforms the input raster plot into a probability with the following formula:
\begin{equation}\label{eq:mlr}
    \forall x, y, \timev, \forall \class \in [1, N_\class], \; \;
Pr(k=\class \; \vert \;  x, y, \timev) =
\frac {\exp  (\mathcal{C}_\class(x, y, \timev) +\bias_\class) }{Z(x, y, \timev)}
\end{equation} 
where $Z(x, y, \timev)=\sum_{\class \in [1, N_\class]} \exp  (\mathcal{C}_\class(x, y, \timev) +\bias_\class)$ is the partition function and  $\bias_\class$ is the bias linked to class $\class$. 
In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike. The spiking output of the model corresponds to an event with the highest probability class.

@@

We would like to highlight a few innovations in the contributions that are presented in this paper. First, whereas~\citep{ghosh_spatio-temporal_2019,yu_stsc-snn_2022} use a correlation-based heuristic, which we observed to be less efficient, the generic heterogeneous model is formalized from first principles for optimal detection of the event-based spatio-temporal motifs. Moreover, in comparison to HOTS~\citep{lagorce_hots_2017}, the weights are explicable as they directly inform on the logit (inverse sigmoid of the probability) of detection of each spatio-temporal spike motif. Another novelty is that the model learns the weights and the delays simultaneously. For example, the polychronization model~\citep{izhikevich_polychronization_2006} learns only the weights using STDP, while the delays are randomly drawn and their values are frozen during learning. In addition, the model is evaluated on a realistic task, while models such as the tempotron are tested on simplified toy problems~\citep{gutig_tempotron_2006}. Another major contribution is to provide a model that is suitable for learning any kind of spatio-temporal spiking motifs and that can be trained in a supervised manner by providing a dataset of supervision pairs. Instead of relying on a careful description of the physical rules governing a task, e.g. the luminance conservation principle for motion detection~\citep{benosman_asynchronous_2012, dardelet_event-by-event_2021}, this allows a more flexible definition of the model using this properly labeled dataset.
%
\subsection{Main limits}
%%%-----------------------------------------------------------------
We have identified a number of limitations of our model, which we will now discuss in detail. First, the entire framework is based on a discrete binning of time, which is not compatible with the continuous nature of biological time. We used this binning to efficiently implement the framework on conventional hardware, especially GPUs, to be able to use fast three-dimensional convolutions. We have tested the effect of the width of the time bin and shown that it has essentially no effect on the results presented in this paper. This is consistent with the relative robustness of other event-based frameworks such as HOTS~\citep{lagorce_hots_2017}, where accuracy was unaffected when the input spikes were subjected to noisy perturbations up to $1~\ms$~\citep{grimaldi_robust_2022}. This suggests the possibility of analytically including a precision term in the temporal value of the input spikes. This mechanism may be implemented by the filtering implemented by the synaptic time constant of about $5~\ms$. Furthermore, it is possible to circumvent the need for time discretization by the use of a purely event-based scheme. In fact, it is not necessary to compute the voltage traces between two spikes~\citep{hanuschkin_general_2010}. Thus, it is possible to define a purely event-based framework. Such an architecture could provide promising computational speedups. % \note{pourrait parler du poster bernstein? de nadafian?}
